High geomagnetic field intensity in the late Mesoproterozoic recorded by Midcontinent Rift anorthosite xenoliths

### 2021-10-12 manuscript/Manuscript.tex
update on manuscript

\begin{keypoints}
\item New \textit{ca.} 1092 Ma Midcontinent Rift anorthosite record high paleointensity values
\item High VADM of the geomagnetic field is inconsistent with previous interpretation of Proterozoic long term decay
\item A strong geodynamo like today likely persisted for about 25 myr during the Midcontinent Rift development
\end{keypoints}

@@

Paleomagnetic data from ancient rocks are one of the few types of observational data that have the potential to record the evolution of Earth’s core and can be used to test the predictions from numerical simulations. Some Precambrian absolute paleointensity records from the PINT database (\url{http://earth.liv.ac.uk/pint/}) show evidence of an existence of an appreciable geomagnetic field prior to the Ediacaran. On the other hand, studies have found very low axial dipole moment (VDM) estimates from rocks formed in the late Ediacaran \cite{Bono2019a, Shcherbakova2019a, Thallner2021b}, followed by a recovery near the Ediacaran-Cambrian boundary \cite{Thallner2021a}. Other studies focusing on paleomagnetic directions have also found that the late Ediacaran might have been a period of time when the geomagnetic field were experiencing frequent reversals and waning of dipole dominance \cite{Bono2015a, Kodama2020a}. It has been interpreted that these findings are consistent with the results of geodynamo models using high core thermal conductivity values that suggest the inner core nucleation age being near the Ediacaran-Cambrian boundary \cite{Driscoll2016a, Davies2021a}. \citeA{Bono2019a} based on a polynomial fit through filtered Precambrian paleointensity database to suggest the selected observational records show that the geodynamo power progressively decreased through the Proterozoic (Figure \ref{fig:MCR_VADM}). That study also recovered very low paleointensity values from \textit{ca.} 565 Ma single silicate crystals and interpreted them to be reflecting that the geodynamo reached its lowest near the Ediacaran-Cambrian boundary. New low paleointensity estimates derived from the \textit{ca.} 720 Ma Franklin Large Igneous Province are interpreted to support the existence of such decay trend in the Neoproterozoic \cite{Lloyd2021a}.

However, previous paleointensity data from intrusive and extrusive rock units in the Mesoproterozoic Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record Earth's virtual axial dipole moment (VADM) values similar to that of today, which are higher than the prediction of the long term decay curve (Figure \ref{fig:MCR_VADM}). \citeA{Pesonen1983b} obtained an average virtual dipole moment of 93 ZAm$^2$ after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using both extrusive and intrusive rocks, including the \textit{ca.} 1108-1105 Ma Osler Volcanics, the Maimainse Point volcanics, intrusive Logan Sills, Thunder Bay dikes, the Baraga and Marquette dikes, as well as lava flow of the \textit{ca.} 1096 Ma North Shore Volcanic Group. \citeA{Kulakov2013a} obtained an average VADM of 59 $\pm$ 11 ZAm$^2$ using the \textit{ca.} 1085 Ma Lake Shore Traps volcanics and suggest that the present-day-like geomagnetic field strength likely existed through the late stage of Midcontinent Rift magmatism \cite{Vervoort2007a, Miller2013a}. However, non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained by \citeA{Pesonen1983b, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. \citeA{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas \citeA{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by \citeA{Sprain2018a} who applied stricter paleointensity selection criteria on new paleointensity results and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. A calculated mean VADM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years \cite{Sprain2018a}. This high VADM value estimate from the Midcontinent Rift volcanics may be incompatible with the hypothesis of a ``Proterozoic dipole low" \cite{Biggin2009a}.

Although it is typical of paleointensity experiments to suffer from low success rates, bulk plagioclase-rich rocks \cite{Selkin2000a} and single plagioclase crystals \cite{Tarduno2005a} are found to be able to produce high quality paleointensity results. It has been suggested that plagioclase crystals can protect magnetic minerals that exsolved within them against post-formation weathering and thermochemical alteration during paleointensity experiments \cite{Tarduno2005a}. In this study, we present new paleointensity results from ``IZZI" modified absolute paleointensity experiments \cite{Yu2004a} on a suite of bulk samples from \textit{ca.} 1092 Ma anorthosite xenoliths and their Beaver River diabase host rocks of the Beaver Bay Complex within the Midcontinent Rift. Paleointensity experiments on the Beaver River anorthosite xenoliths have a very high success rate. The consistent specimen- and site-level paleointensity results reveal that the magnetite-dominated remanence of the anorthosite specimens has low thermal remanent magnetization (TRM) anisotropy and linear TRM acquisition. Magnetic imaging shows that the anorthosite specimens to have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. With more single-slope, high quality paleointensity data, we seek to compare paleointensity records from the Midcontinent Rift intrusives with those from the extrusives; to test the hypothesis that the Earth's geomagnetic dipole strength remained at high values throughout the Midcontinent Rift; and to further enrich the reliable paleointensity records during the $\sim$25 myr of Midcontinent Rift magmatic activity.

@@

Despite that the anorthosite xenoliths of the Beaver River diabase were entrained as xenoliths and emplaced only $\sim$4 myr after the formation of the Anorthositic Series rocks of the Duluth Complex when massive plagioclase mushes were generated, $^{206}$Pb/$^{238}$U zircon geochronology data and zircon rare earth element diffusion modeling results suggest it is the most probable that the Beaver River diabase anorthosite xenoliths are entrained cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. Moreover, the anorthosite xenoliths of the \textit{ca.} 1092 Ma Beaver River diabase has distinct mineralogy from that of the Anorthositic Series rocks of the Duluth Complex. The dominantly monomineralic anorthosite xenoliths have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. In addition, petrographic imaging support that the plagioclase from the Beaver River anorthosite formed under a magmatic conditions different from the Duluth Complex anorthositic series magma. While the plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture (Figure \ref{Petro_QDM}A), the more calcium-rich plagioclase of the Beaver River anorthosite often develop granoblastic texture characterized by equigranular plagioclase crystals (Figure \ref{Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial vermicular titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex Anorthositic Series rocks \cite{Miller1990a}.

@@

\subsection*{Late Mesoproterozoic geodynamo }
To best characterize the geomagnetic axial dipole field intensity during a certain time period requires that a paleointensity dataset to cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged and the geocentric axial dipole hypothesis (GAD) can be assumed. Thermal modeling results from \citeA{Zhang2021b} suggest that the anorthosite xenoliths were likely heated to tholeiitic magma temperature and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a few thousand years. In addition, the closely located sites AX12 and AX16 have very similar paleointensity estimates that are distinct from those of sites AX11 and AX13 which were emplaced $\sim$10 km away (Figure \ref{fig:Geologic_map}). The consistency within small regions but variations between regions could be the result of differential cooling history due to heterogeneity of the timing and depths of the Beaver River diabase intrusions, causing the anorthosites to capture paleosecular variations \textit{ca.} 1092 Ma. Overall, none of the anorthosite xenoliths record a V(A)DM value smaller than 40 ZAm$^2$ (Figure \ref{fig:PINT_cooling_corrected}).

Such a high estimate for the late Mesoproterozoic may invoke more careful examination of the geodynamo evolution through the Proterozoic and the timing for the inner core nucleation. Paleomagnetic observational records near the Ediacaran-Cambrian boundary have been interpreted to be consistent with a trend of decreasing axial dipole strength toward a very low value \cite{Lloyd2021a, Bono2019a}, followed by a rapid recovery (a few tens of million years) toward higher values soon after \cite{Thallner2021a}. Such a transition near this time can be consistent with the dynamo numerical simulations which adopt a rather high thermal conductivity of the core \cite{Davies2021a}. In particular, some models have suggested that the kick-start of rigorous compositional convection between the inner core and the outer core soon after the onset of inner core nucleation can result in almost an order of magnitude increase in Earth's axial dipole moment \cite{Davies2021a}.

Although more recent data

using paleomagnetic observational records is guided by the predictions from numerical simulations that a cooling liquid core with only thermal convection power to drive a dynamo would experience a progressive decay in its magnetic dipole field strength, and experience a rapid recover after the onset of inner core nucleation \cite{Davies2021a}. Previous studies has pointed to the Ediacaran-Cambrian boundary for such a weakening field behavior and recent studies has claim to found a rapid increase after the single crystal time which fits in the numerical models. However, such a recover may not be unique given the strong geodynamo estimate from the 1.1 Ga Midcontinent Rift. One could point to the increase from low values from the ca. 1140 Ma abitibi dikes to the ca. 1092 Ma anorthosites to be the result of recovery; another line of disagreement is between the Bangemall sills and the low field from Sept Ils. MCR being an anomaly? or the time scale we are looking at is too long?

Regardless these several lines of increase/decrease trend needs further explanations/considerations.

use a combined method of direction and intensity research for further examinations. Re-evaluation of the spatial and temporal variability captured by the existing Precambrian database. More data are needed for the Precambrian in order to better characterize the geodynamo evolution.

\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{PINT_compilation.pdf}
\centering
\caption{\footnotesize{Compilation of calculated virtual (axial) dipole moment (VADM) values from the PINT database (v.2015.05; \url{http://earth.liv.ac.uk/ pint/}; \citeA{Biggin2015a}) with additional Mesoproterozoic data from \citeA{Sprain2018a, Kodama2019a}, as well as those from the Neoproterozoic \cite{Lloyd2021a, Lloyd2021b, Thallner2021a, Thallner2021b}. Paleointensity estimates from \citeA{Pesonen1983a} and \citeA{Kulakov2013a} are not included in the compilation due to the specimen-level double-slope behavior as discussed in the text. Overall the anorthosite xenoliths from this study record a high Mesoproterozoic field exceeding the value projected by the second order polynomial curve from \citeA{Bono2019a} which is based on an interpretation of there being a progressive decay of the geodynamo through the Proterozoic.}}
\label{fig:MCR_VADM}
\end{figure}

The calculated paleomagnetic VADM values are shown in Figure \ref{} in context of the compilation of all Midcontinent Rift results. highlight the high quality of the results and the consistency of MCR rocks' high geomagnetic field values even considering the uncertainties of previous data. Supporting that the Mesoproterozoic likely experienced a period of persistent high geomagnetic strength in addition to the protracted magmatism of in the formation of the massive intracontinental rifting.

@@

% that this dataset also could have suffered from problem of averaging out secular variations, given that the mean pmag directions deviate from what is expected from the sythesized APWP, nevertheless the variations in estimate absolute paleointensity lead us to interpret that albeit close, the they intensity data were acquired at distinctly different times that are sufficiently far apart such that the surface geomagnetic field had already changed.

@@

\section*{Conclusion}
high success rate comparing to volcanics (more than half success sites), producing good consistent site level results, and across sites report present-day like field like today.

AF pre-treatment is helpful in that it increase the specimens/sites passing the selecting criteria, suggesting that the method likely helped improve the MD ferromagnetic carriers.

Rock mag and petrographic data might suggest differences in terms of the value of peak field on coercivity spectra between those passed and those didn't.

IRM measured before and after an acid dunk in --- concentration acid on specimens after IZZI experiment show a difference (significant drop) maybe? between specs that passed and those did not.

geographical association

anisotropy of full TRM experiment on accepted specs show low degree of bulk sample scale full TRM anisotropy, with a range of ---. Cooling rate correction based on \cite{Halgedahl1980a} suggest a lab over estimate of B ancient of 35 \%, assuming constant cooling rate in natural condition. The cooling rate corrected field is $\sim$40 $\mu$T which is similar to today's field strength at similar latitude.

>>>>

\begin{keypoints}
\item New \textit{ca.} 1092 Ma Midcontinent Rift anorthosite record high paleointensity values
\item High VADM of the geomagnetic field is inconsistent with previous interpretation of Proterozoic long term decay
\item A strong geodynamo like today likely persisted for about 25 Myr during the Midcontinent Rift development
\end{keypoints}

@@

Paleomagnetic data from ancient rocks are one of the few types of observational data that have the potential to record the evolution of Earth’s core and can be used to test the predictions from numerical simulations. Some Precambrian absolute paleointensity records from the PINT database (\url{http://earth.liv.ac.uk/pint/}) show evidence of an existence of an appreciable geomagnetic field prior to the Ediacaran. On the other hand, studies have found very low axial dipole moment (VDM) estimates from rocks formed in the late Ediacaran \cite{Bono2019a, Shcherbakova2019a, Thallner2021b}, followed by a recovery near the Ediacaran-Cambrian boundary \cite{Thallner2021a}. Other studies focusing on paleomagnetic directions have also found that the late Ediacaran might have been a period of time when the geomagnetic field were experiencing frequent reversals and waning of dipole dominance \cite{Bono2015a, Kodama2020a}. It has been interpreted that these findings are consistent with the results of geodynamo models using high core thermal conductivity values that suggest the inner core nucleation age being near the Ediacaran-Cambrian boundary \cite{Driscoll2016a, Davies2021a}. \citeA{Bono2019a} based on a polynomial fit through filtered Precambrian paleointensity database to suggest the selected observational records show that the geodynamo power progressively decreased through the Proterozoic (Figure \ref{fig:PINT_compilation}). That study also recovered very low paleointensity values from \textit{ca.} 565 Ma single silicate crystals and interpreted them to be reflecting that the geodynamo reached its lowest near the Ediacaran-Cambrian boundary. New low paleointensity estimates derived from the \textit{ca.} 720 Ma Franklin Large Igneous Province are interpreted to support the existence of such decay trend in the Neoproterozoic \cite{Lloyd2021a}.

However, previous paleointensity data from intrusive and extrusive rock units in the Mesoproterozoic Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record Earth's virtual axial dipole moment (VADM) values similar to that of today, which are higher than the prediction of the long term decay curve (Figure \ref{fig:PINT_compilation}). \citeA{Pesonen1983b} obtained an average virtual dipole moment of 93 ZAm$^2$ after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using both extrusive and intrusive rocks, including the \textit{ca.} 1108-1105 Ma Osler Volcanics, the Maimainse Point volcanics, intrusive Logan Sills, Thunder Bay dikes, the Baraga and Marquette dikes, as well as lava flow of the \textit{ca.} 1096 Ma North Shore Volcanic Group. \citeA{Kulakov2013a} obtained an average VADM of 59 $\pm$ 11 ZAm$^2$ using the \textit{ca.} 1085 Ma Lake Shore Traps volcanics and suggest that the present-day-like geomagnetic field strength likely existed through the late stage of Midcontinent Rift magmatism \cite{Vervoort2007a, Miller2013a}. However, non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained by \citeA{Pesonen1983b, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. \citeA{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas \citeA{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by \citeA{Sprain2018a} who applied stricter paleointensity selection criteria on new paleointensity results and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. A calculated mean VADM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years \cite{Sprain2018a}. This high VADM value estimate from the Midcontinent Rift volcanics may be incompatible with the hypothesis of a ``Proterozoic dipole low" \cite{Biggin2009a}.

Although it is typical of paleointensity experiments to suffer from low success rates, bulk plagioclase-rich rocks \cite{Selkin2000a} and single plagioclase crystals \cite{Tarduno2005a} are found to be able to produce high quality paleointensity results. It has been suggested that plagioclase crystals can protect magnetic minerals that exsolved within them against post-formation weathering and thermochemical alteration during paleointensity experiments \cite{Tarduno2005a}. In this study, we present new paleointensity results from ``IZZI" modified absolute paleointensity experiments \cite{Yu2004a} on a suite of bulk samples from \textit{ca.} 1092 Ma anorthosite xenoliths and their Beaver River diabase host rocks of the Beaver Bay Complex within the Midcontinent Rift. Paleointensity experiments on the Beaver River anorthosite xenoliths have a very high success rate. The consistent specimen- and site-level paleointensity results reveal that the magnetite-dominated remanence of the anorthosite specimens has low thermal remanent magnetization (TRM) anisotropy and linear TRM acquisition. Magnetic imaging shows that the anorthosite specimens to have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. With more single-slope, high quality paleointensity data, we seek to compare paleointensity records from the Midcontinent Rift intrusives with those from the extrusives; to test the hypothesis that the Earth's geomagnetic dipole strength remained at high values throughout the Midcontinent Rift; and to further enrich the reliable paleointensity records during the $\sim$25 Myr of Midcontinent Rift magmatic activity.

@@

Despite that the anorthosite xenoliths of the Beaver River diabase were entrained as xenoliths and emplaced only $\sim$4 Myr after the formation of the Anorthositic Series rocks of the Duluth Complex when massive plagioclase mushes were generated, $^{206}$Pb/$^{238}$U zircon geochronology data and zircon rare earth element diffusion modeling results suggest it is the most probable that the Beaver River diabase anorthosite xenoliths are entrained cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. Moreover, the anorthosite xenoliths of the \textit{ca.} 1092 Ma Beaver River diabase has distinct mineralogy from that of the Anorthositic Series rocks of the Duluth Complex. The dominantly monomineralic anorthosite xenoliths have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. In addition, petrographic imaging support that the plagioclase from the Beaver River anorthosite formed under a magmatic conditions different from the Duluth Complex anorthositic series magma. While the plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture (Figure \ref{Petro_QDM}A), the more calcium-rich plagioclase of the Beaver River anorthosite often develop granoblastic texture characterized by equigranular plagioclase crystals (Figure \ref{Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial vermicular titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex Anorthositic Series rocks \cite{Miller1990a}.

@@

\subsection*{Late Mesoproterozoic geodynamo }
To best characterize the geomagnetic axial dipole field intensity during a certain time period requires that a paleointensity dataset to cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged and the geocentric axial dipole hypothesis (GAD) can be assumed. Thermal modeling results from \citeA{Zhang2021b} suggest that the anorthosite xenoliths were likely heated to tholeiitic magma temperature and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a few thousand years. In addition, the closely located sites AX12 and AX16 have very similar paleointensity estimates that are distinct from those of sites AX11 and AX13 which were emplaced $\sim$10 km away (Figure \ref{fig:Geologic_map}). The consistency within small regions but variations between regions could be the result of differential cooling history due to heterogeneity of the timing and depths of the Beaver River diabase intrusions, causing the anorthosites to capture paleosecular variations \textit{ca.} 1092 Ma. Overall, none of the anorthosite xenoliths record a V(A)DM value smaller than 40 ZAm$^2$ (Figure \ref{fig:PINT_compilation}).

Paleomagnetic observational records near the Ediacaran-Cambrian boundary have been interpreted to be consistent with a trend of decreasing axial dipole strength toward a very low value \cite{Lloyd2021a, Bono2019a}, followed by a rapid recovery (a few tens of million years) toward higher values soon after \cite{Thallner2021a}. Such a decrease-increase transition can be consistent with the dynamo numerical simulations which adopt a rather high thermal conductivity of the core and projects an inner core birth near this time \cite{Gomi2013a, Davies2015a, Labrosse2015a, Ohta2016a}. In particular, some models have suggested that the kick-start of rigorous compositional convection between the inner core and the outer core soon after the onset of inner core nucleation can result in almost an order of magnitude increase in Earth's axial dipole moment \cite[e.g.]{Davies2021a}. The enriching observational records from the Ediacaran and Cambrian have been suggested to indicate that a geodynamo driven solely by thermal convection was waning to its brink of collapse near the end of Ediacaran \cite{Bono2019a, Thallner2021b} and was recovering to ward modern day strengths soon after \cite{Thallner2021a}.

However, the updated paleointensity records from the Midcontinent Rift together with previous results from other Mesoproterozoic rocks suggest that the a dramatic increase in geomagnetic dipole moment similar in scale to that modeled for after the onset of inner core nucleation may not be unique to the Edicaran-Cambrian boundary before 500 Ma (Figure \ref{fig:PINT_compilation}). The \textit{ca.} 1140 Ma Abitibi dikes \cite{Macouin2003a} record an average virtual dipole moment of $\sim$13 ZAm$^2$, similar to the low values of $\sim$7 ZAm$^2$ obtained from the single silicate crystals of the \textit{ca.} 565 Ma Sept-\hat{I}les intrusions \cite{Bono2019a}. However, the relatively high virtual dipole moments of $\sim$72 ZAm$^2$ from the \textit{ca.} 1105 Ma Osler Volcanic Group \cite{Sprain2018a} and $\sim$82 ZAm$^2$ from the \textit{ca.} 1092 Ma Beaver River anorthosite xenoliths (Figure \ref{fig:PINT_compilation} are about six times the values from the \textit{ca.} 1140 Ma Abitibi dikes. In addition, the increase in absolute dipole moment estimates recorded by the Midcontinent Rift rocks are much higher than that in the Ediacaran (Figure \ref{fig:PINT_compilation}).

Furthermore, the high geomagnetic field estimates from the Midcontinent Rift volcanics and intrusives are not the only suite of rocks that bear records at odds with the hypothesis of a progressive decay trend during the Proterozoic \cite{Biggin2009a, Bono2019a}. The $\sim$60 ZAm$^2$ dipole moment estimate from the \textit{ca.} 755 Ma Mundine Well dikes are discordant with the long-term decrease trend \cite{Bono2019a} and is about three times the estimates from the \textit{ca.} 720 Ma dolerite dikes of the Franklin Large Igneous Province \cite{Lloyd2021a}. Similarly, an apparent rapid decrease in geomagnetic dipole moment from \textit{ca.} 1092 Ma to \textit{ca.} 1070 Ma can be evidenced by the contrasting high values from the Midcontinent Rift anorthosite xenoliths and the low values from the Mundine Well dikes (Figure \ref{PINT_Compilation}). However, a sparse record in a short time period may not sufficiently average out paleosecular variations. Given that a five-fold fluctuation of the geomagnetic dipole moment exist in the paleointensity records for the past 10 million years (5 and 95 percentiles), when we know a modern-day-like geodynamo likely powered by both thermal and compositional convection existed, we cannot conclude on whether an apparent rise or decay within a narrow time period ($\sim$35 Ma) during the Proterozoic is associated with transitions in the core fluid regime or is a result of short-term variations of the geomagnetic field that may be common through much of the evolution of Earth's core. In addition, the observations of geomagnetic dipole moment fluctuations in more recent time scales may suggest that evaluations of long-term evolution of the Precambrian paleointensity records should be accompanied by development of higher-resolution data in more short-term scales.

Overall, the existing absolute paleointensity records in the PINT database remains sparse in terms of spatial and temporal coverage as compared to more geologically recent times. (Figure \ref{PINT_compilation}). More data are needed in order to better characterize the evolution of Earth's dynamo and thermal evolution.

% 1. whether septils was recording the time averaged field or the low values are from the hyperactive state which is another signal prior to the onset of inner core? indeed it is interesting that no high values from the Ediacaran has been recovered

% 2. where is the recovery, how long was the recovery, how strong was the recovery is at question for he Ediacaran boundary.

% 3. How should one explain the Mundine Well dikes high values?

% 4. How should one explain the Bungemall sills low values in 1070 Ma following the high values of the MCR?

% 5. We need more spatial and temporal coverage of the Precambrian observational data.

% using paleomagnetic observational records is guided by the predictions from numerical simulations that a cooling liquid core with only thermal convection power to drive a dynamo would experience a progressive decay in its magnetic dipole field strength, and experience a rapid recover after the onset of inner core nucleation \cite{Davies2021a}. Previous studies has pointed to the Ediacaran-Cambrian boundary for such a weakening field behavior and recent studies has claim to found a rapid increase after the single crystal time which fits in the numerical models. However, such a recover may not be unique given the strong geodynamo estimate from the 1.1 Ga Midcontinent Rift. One could point to the increase from low values from the ca. 1140 Ma abitibi dikes to the ca. 1092 Ma anorthosites to be the result of recovery; another line of disagreement is between the Bangemall sills and the low field from Sept Ils. MCR being an anomaly? or the time scale we are looking at is too long?

% use a combined method of direction and intensity research for further examinations. Re-evaluation of the spatial and temporal variability captured by the existing Precambrian database. More data are needed for the Precambrian in order to better characterize the geodynamo evolution.

\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{PINT_compilation.pdf}
\centering
\caption{\footnotesize{Compilation of calculated virtual (axial) dipole moment (VADM) values from the PINT database (v.2015.05; \url{http://earth.liv.ac.uk/ pint/}; \citeA{Biggin2015a}) with additional Mesoproterozoic data from \citeA{Sprain2018a, Kodama2019a}, as well as those from the Neoproterozoic \cite{Lloyd2021a, Lloyd2021b, Thallner2021a, Thallner2021b}. Paleointensity estimates from \citeA{Pesonen1983a} and \citeA{Kulakov2013a} are not included in the compilation due to the specimen-level double-slope behavior as discussed in the text. Overall the anorthosite xenoliths from this study record a high Mesoproterozoic field exceeding the value projected by the second order polynomial curve from \citeA{Bono2019a} which is based on an interpretation of there being a progressive decay of the geodynamo through the Proterozoic.}}
\label{fig:PINT_compilation}
\end{figure}

@@

@@

% Regardless, a strong geomagnetic field has been recorded by the \textit{ca.} 1092 Ma Beaver River anorthosite xenoliths and a strong geodynamo might have persisted through the protracted Midcontinent Rift magmatism.

\section*{Conclusion}

The paucity of Precambrian paleointensity estimates, as well as non-ideal results within the existing database, present challenges when seeking to interpret long-term trends in the strength of the geomagnetic field. The Midcontinent Rift of North America presents an opportunity to develop robust paleointensity estimates spanning a $\sim$25 Myr interval of the late Mesoproterozoic. Rock magnetic data support that the \textit{ca.} 1092 Ma anorthosite xenoliths of the Midcontinent Rift diabase can have dominantly high coercivity magnetic mineralogy with minimal anisotropic or nonlinear thermal remanence acquisition, producing high quality paleointensity results. Their overall high estimates of the late Mesoproterozoic virtual dipole moment together with previous Midcontinent Rift records suggest that a strong geomagnetic field like today's might have persisted through the rifting.

The new data in context of the current Precambrian paleointensity database suggest that a strong increase in geomagnetic dipole moment which is thought to be indicative of the onset of compositional convection caused by the inner core nucleation may not be unique during the Precambrian. More spatial and temporal coverage of the observational records is needed for further analyses of the long-term evolution of the geodynamo.
